---
title: "Proportion Test"
author: "Yaamini Venkataraman"
date: "1/15/2019"
output: html_document
---

I will use a proportion test to compare the proportion of genome feature overlaps between differentially methylated loci (DML), differentially methylated regions (DMR), and the gene background.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Obtain session information

```{r}
sessionInfo()
```

# Import data

```{r}
overlapData <- read.csv("2019-01-15-Overlap-Proportions.csv", header = TRUE)
rownames(overlapData) <- overlapData$genomicFeature #Set genomic feature indication and rownames
overlapData <- overlapData[,-1] #Remove genomic feature indication column
head(overlapData) #Confirm import
```

# Conduct chi-squared tests of homogeneity

The null hypothesis is that loci distributions in the genome are the same between different categories.

## Total CpGs vs. all CpGs with 5x coverage enriched by MBD treatment

I want to see if MBD enriched CpGs in certain areas of the genome.

```{r}
totalVersusEnriched <- overlapData[, 1:2] #Subset data for statistical testing
totalVersusEnriched <- totalVersusEnriched[-6, ] #Remove totalLines row
head(totalVersusEnriched) #Confirm changes
```

```{r}
totalVersusEnrichedTest <- chisq.test(totalVersusEnriched) #Conduct a chi-squared test
totalVersusEnrichedTest #The distribution of enriched CpGs from MBD is not different from the distribution of all CG motifs in the C. virginica genome
```

## Enriched CpGs vs. methylated CpGs

Are methylated CpGs more populous in certain genomic regions?

```{r}
enrichedVersusMethylated <- overlapData[, 2:3] #Subset data for statistical testing
enrichedVersusMethylated <- enrichedVersusMethylated[-6, ] #Remove totalLines row
head(enrichedVersusMethylated) #Confirm changes
```

```{r}
enrichedVersusMethylatedTest <- chisq.test(enrichedVersusMethylated) #Conduct a chi-squared test
enrichedVersusMethylatedTest #The distribution of methylated CpGs is significantly different from CpGs enriched by MBD
```

## Enriched CpGs vs. methylated CpGs vs. sparsely methylated CpGs vs. unmethylated CpGs

```{r}
allMethylationCategories <- overlapData[, 2:5] #Subset data for statistical testing
allMethylationCategories <- allMethylationCategories[-6, ] #Remove totalLines row
head(allMethylationCategories) #Confirm changes
```

```{r}
allMethylationCategoriesTest <- chisq.test(allMethylationCategories) #Conduct a chi-squared test
allMethylationCategoriesTest #The distribution of methylated CpGs is significantly different from all CpGs in the C. virginica genome.
```

## Methylated CpGs vs. DML

```{r}
methylatedVersusDML <- overlapData[,c(3,6)] #Subset data for statistical testing
methylatedVersusDML <- methylatedVersusDML[-6,] #Remove totalLines row
head(methylatedVersusDML) #Confirm changes
```

```{r}
methylatedVersusDMLTest <- chisq.test(methylatedVersusDML) #Conduct a chi-squared test
methylatedVersusDMLTest #The distribution of DML is significantly different from methylated CpGs.
```

# Create figures

## Reformat data

```{r}
head(overlapData)
```


```{r}
proportionData <- overlapData #Copy overlap data as a new dataframe
nLength <- length(proportionData$totalCpG) #Count number of rows
for(i in 1:nLength) {
  proportionData[i,] <- (proportionData[i,]/proportionData[6,])*100
} #Divide each column of proportionData by respective totalLines entry. Multiply by 100 and and save the percentage
head(proportionData) #Confirm changes
```

```{r}
proportionData <- proportionData[-6,] #Remove totalLines row
tail(proportionData) #Confirm changes
```

## Enriched CpGs vs. methylated CpGs

```{r}
enrichedVersusMethylatedProportions <- proportionData[,2:3] #Subset data
head(enrichedVersusMethylatedProportions) #Confirm subset
```

```{r}
#pdf("2019-04-10-Total-Versus-Methylated-CpGs.pdf", height = 8.5, width = 11)
par(mar = c(3,5,1,1)) #Change figure boundaries
barplot(t(enrichedVersusMethylatedProportions), 
        beside = TRUE, 
        axes = FALSE, 
        names.arg = c("Exons", "Introns", "TE", "Promoters", "Other")) #Create a grouped barplot (beside = TRUE) using a transposed version of the proportion data. Use axes = FALSE to remove the y-axis and names.arg to set labels on the x-axis.
axis(side = 2, at = seq(0, 45, by = 5), las = 2, col = "grey80")
mtext(side = 2, "Proportion CpGs", line = 3) #Add y-axis label
legend("topright", 
       legend = c("MBD-Enriched", "Methylated"),
       pch = 16,
       col = c("grey20", "grey80"),
       bty = "n") #Place a legend in the top right of the figure with no box
#dev.off()
```

## Enriched CpGs vs. methylated vs. sparsely methylated vs. unmethylated

```{r}
allMethylationCategoriesProportions <- proportionData[,2:5] #Subset data
head(allMethylationCategoriesProportions) #Confirm subset
```

```{r}
#pdf("2019-04-10-All-Methylation-Categories.pdf", height = 8.5, width = 11)
par(mar = c(3,5,1,1)) #Change figure boundaries
barplot(t(allMethylationCategoriesProportions), 
        beside = TRUE, 
        axes = FALSE, 
        names.arg = c("Exons", "Introns", "TE", "Promoters", "Other")) #Create a grouped barplot (beside = TRUE) using a transposed version of the proportion data. Use axes = FALSE to remove the y-axis and names.arg to set labels on the x-axis.
axis(side = 2, at = seq(0, 45, by = 5), las = 2, col = "grey80")
mtext(side = 2, "Proportion CpGs", line = 3) #Add y-axis label
legend("topright", 
       legend = c("MBD-Enriched", "Methylated", "Sparsely Methylated", "Unmethylated"),
       pch = 16,
       col = c("grey20", "grey40", "grey60", "grey80"),
       bty = "n") #Place a legend in the top right of the figure with no box
#dev.off()
```

## Methylated CpGs vs. DML

```{r}
methylatedVersusDMLProportions <- proportionData[,c(3,6)] #Subset data
head(methylatedVersusDMLProportions) #Confirm subset
```

```{r}
#pdf("2019-04-10-Methylated-CpGs-Versus-DML.pdf", height = 8.5, width = 11)
par(mar = c(3,5,1,1)) #Change figure boundaries
barplot(t(methylatedVersusDMLProportions), 
        beside = TRUE, 
        axes = FALSE, 
        names.arg = c("Exons", "Introns", "TE", "Promoters", "Other")) #Create a grouped barplot (beside = TRUE) using a transposed version of the proportion data. Use axes = FALSE to remove the y-axis and names.arg to set labels on the x-axis.
axis(side = 2, at = seq(0, 60, by = 5), las = 2, col = "grey80")
mtext(side = 2, "Proportion CpGs", line = 3) #Add y-axis label
legend("topright", 
       legend = c("Methylated CpGs", "DML"),
       pch = 16,
       col = c("grey20", "grey80"),
       bty = "n") #Place a legend in the top right of the figure with no box
#dev.off()
```