---
title: "Gene Enrichment Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script, I'll isolate Uniprot codes needed to perform a gene enrichment. My ultimate goal is to identify enriched genes for mRNA regions that overlap with DMRs and DMLs. I will isolate Uniprot codes for the DMR-mRNA overlap, and for the gene background. The gene background will serve as my background for enrichment analyses, instead of the full transcriptome.

# Install packages

```{r}
require(dplyr)
```

# Session information

```{r}
sessionInfo()
```

# Obtain Uniprot codes

Before I do any sort of gene enrichment analysis, I need to get Uniprot codes for all genes. To do this, I will use `blastx`.

## Make a Uniprot Database

The Uniprot database was downloaded from this link as a .fasta file: http://www.uniprot.org/uniprot/?query=&fil=reviewed%3Ayes&columns=id%2Centry%20name%2Creviewed%2Cprotein%20names%2Cgenes%2Corganism%2Clength%2Cgo(biological%20process)%2Cgo-id%2Ccomment(PATHWAY)%2Cdatabase(UniPathway)%2Cdatabase(CDD)%2Cdatabase(Pfam.

```{bash}
head uniprot-filtered-reviewed.fasta #Check database creation
```

```{bash}
/Users/Shared/Apps/ncbi-blast-2.2.29+/bin/makeblastdb -help #Get the help menu
```

The first step involves taking the uniprot fasta file and making a new database. To do this, I will use the following code:

1. Path to `makeblastdb` to create the database needed
2. `in` to select the input file for the database
3. `dbtype` specifies the type of file which in our case, is a protein file

```{bash}
/Users/Shared/Apps/ncbi-blast-2.2.29+/bin/makeblastdb \
-in uniprot-filtered-reviewed.fasta \
-dbtype 'prot'
```

```{bash}
/Users/Shared/Apps/ncbi-blast-2.2.29+/bin/blastx -help
```

Now I can perform a `blastx` to compare my query, the *C. virginica* transcripts, to our protein database.

## Perform `blastx`

1. Path to `blastx`
2. `-query` provides the file we want to blast (*C. virginica* transcripts)
3.`-db` specifies database created in the previous step
4. `-outfmt` specifies the type of output file. I will use 6, a tabular file
5. `-out <filename>` will allow me to save the output as a new file
6. `-num_threads 4` uses 4 CPUs in the BLAST search
7. `-max_target_seqs 1` keeps only one aligned sequence (i.e. the best match)

```{bash}
pwd #Obtain present working directory
```

```{bash}
/Users/Shared/Apps/ncbi-blast-2.2.29+/bin/blastx \
-query 2018-09-06-Virginica-transcripts.fna \
-db uniprot-filtered-reviewed.fasta \
-outfmt 6 \
-out 2018-09-06-Transcript-Uniprot-blastx.txt \
-num_threads 4 \
-max_target_seqs 1
```

```{bash}
head 2018-09-11-Transcript-Uniprot-blastx-codeIsolated.txt #Check file format of Uniprot codes. I'm using a file with the Uniprot code already isolated so I don't have to unfold any columns after merging files.
```

```{r}
blastResults <- read.delim("2018-09-11-Transcript-Uniprot-blastx-codeIsolated.txt", header = FALSE, sep = "\t", dec = ".") #Import blast results
head(blastResults) #Confirm import
colnames(blastResults) <- c("Genbank", "V2", "Uniprot", "V4", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
head(blastResults) #Confirm column headers
```

# Master gene list

I want to create a master list with the following columns:

- chr
- start
- end
- p-value
- q-value
- meth.diff
- dist-to-closest-gene
- gene-chr
- gene-start
- gene-end
- gene-ID
- mRNA-start
- mRNA-end
- genbank
- uniprot-ID
- e-value
- product

I can start with my DML csv file, then merge that with the information from `closestBed`. Then I'll merge that document with my gene track to get gene IDs. I'll merge the resultant file with the mRNA track by gene ID to get Genbank IDs. Finally, I'll merge that document with the Uniprot codes by matching Genbank IDs.

## Merge DML list with `closestBed` list

```{r}
allDML <- read.csv("../2018-10-25-MethylKit/2018-10-25-Loci-Analysis/2019-04-05-Differentially-Methylated-Loci-Filtered-Destrand-50-Cov5.csv", header = TRUE) #Import DML list
allDML <- allDML[,-1] #Remove first column
head(allDML) #Confirm import
```

```{r}
closestDML <- read.delim("../2018-11-01-DML-and-DMR-Analysis/2019-05-29-Flanking-Analysis/2019-05-29-Genes-Closest-NoOverlap-DMLs.txt", header = FALSE) #Import DML with closestBed information
colnames(closestDML) <- c("chr", "start", "end", "meth.diff", "gene-chr", "gene-start", "gene-end", "dist-to-closest-gene") #Add column names
head(closestDML) #Confirm column names. DML downstream of genes have positive distances. There may be more lines than DML becuase there are some DML that overlap multiple gene features, or have ties for closest gene features.
```

```{r}
#Add 1 to the start position and subtract 1 from the end position to get DML positions compatible with the DML list
closestDML$start <- closestDML$start + 1
closestDML$end <- closestDML$end -1
head(closestDML) #Confirm changes
```

```{r}
allDMLclosestBed <- merge(x = closestDML, y = allDML, by = "start", sort = FALSE) #Merge closest information and DML list. Do not resort the output.
head(allDMLclosestBed) #Confirm merge
```

```{r}
allDMLclosestBed <- allDMLclosestBed[, c(2, 1, 3, 4, 11, 12, 13, 14, 8, 5, 6, 7)] #Reorganize columns
head(allDMLclosestBed) #Confirm reorganization
```

## Merge with gene track

### Import full gene list

```{r}
geneList <- read.delim("../2019-05-13-Generating-Genome-Feature-Tracks/C_virginica-3.0_Gnomon_gene_sorted_yrv.gff3", header = FALSE) #Import gene gff3
colnames(geneList) <- c("gene-chr", "Gnomon", "gene", "gene-start", "gene-end", "V6", "V7", "V8", "V9") #Rename columns
head(geneList) #Confirm import
```

```{r}
geneList <- geneList[,-c(6:8)] #Remove columns 6-8
head(geneList) #Confirm changes
```

### Isolate gene description from full list

```{bash}
cut -f9 ../2019-05-13-Generating-Genome-Feature-Tracks/C_virginica-3.0_Gnomon_gene_sorted_yrv.gff3 \
| awk  '{gsub(";","\t",$0); print;}' \
> 2019-06-20-Gene-Descriptions.txt #Isolate gene description column and convert ";" to "\t". Save the output as a new file
```

```{bash}
head 2019-06-20-Gene-Descriptions.txt #Confirm file creation. 
```

Before I import this new data set into R, I need to count the number of columns. I can do this with `awk`.

```{bash}
awk '{print NF}' 2019-06-20-Gene-Descriptions.txt > 2019-06-20-Gene-Description-Column-Numbers.txt
```

### Import isolated gene descriptions

```{r}
geneDescriptionColumns <- read.delim("2019-06-20-Gene-Description-Column-Numbers.txt", header = FALSE) #Import data with number of columns for each row
max(geneDescriptionColumns) #Max number of columns. I'll need to import my gene description dataset with 9 columns per row.
```

```{r}
geneDescriptions <- read.delim2("2019-06-20-Gene-Descriptions.txt", header = FALSE, col.names = c("gene-number", "gene-ID", "V3", "V4", "V5", "V6", "V7", "V8", "V9")) #Import document with gene descrptions. Name 9 columns so all rows are imported with 9 columns
head(geneDescriptions) #Confirm import
```


```{r}
geneID <- geneDescriptions$gene.ID #Isolate gene ID column
geneID <- gsub("Dbxref=GeneID:", "", geneID) #Remove extraneous characters
head(geneID) #Confirm changes
```

```{r}
geneList <- data.frame(geneList, geneID) #Replace geneList dataframe with geneList and geneID
geneList <- geneList[,-6] #Remove extra column
colnames(geneList) <- c("gene-chr", "Gnomon", "gene", "gene-start", "gene-end", "gene-ID") #Rename columns to match previous naming conventions
head(geneList) #Confirm merge
```

### Merge gene track with `allDMLclosestBed`

```{r}
allDMLclosestBedgeneList <- merge(x = allDMLclosestBed, y = geneList, by = c("gene-start", "gene-end"), sort = FALSE) #Merge dataframes by gene start and end positions
head(allDMLclosestBedgeneList) #Confirm merge
```

```{r}
allDMLclosestBedgeneList <- allDMLclosestBedgeneList[,c(3:12, 14:15, 1:2, 16)] #Reorganize columns
head(allDMLclosestBedgeneList) #Confirm reorganization
```

## Merge with mRNA track

### Import full mRNA track

```{r}
mRNAList <- read.delim("../2019-05-13-Generating-Genome-Feature-Tracks/C_virginica-3.0_Gnomon_mRNA_yrv.gff3", header = FALSE) #Import mRNA track
mRNAList <- mRNAList[,-c(6,8)] #Remove extra columns
colnames(mRNAList) <- c("mRNA-chr", "Gnomon", "mRNA", "mRNA-start", "mRNA-end", "mRNA-strand", "V9") #Rename columns
head(mRNAList) #Confirm import
```

### Isolate mRNA description from full list

```{bash}
cut -f9 ../2019-05-13-Generating-Genome-Feature-Tracks/C_virginica-3.0_Gnomon_mRNA_yrv.gff3 \
| awk  '{gsub(";","\t",$0); print;}' \
| awk  '{gsub(",","\t",$0); print;}' \
> 2019-06-20-mRNA-Descriptions.txt #Isolate mRNA description column and convert ";" and "," to "\t". Save the output as a new file
```

```{bash}
head 2019-06-20-mRNA-Descriptions.txt #Confirm file creation. 
```

Before I import this new data set into R, I need to count the number of columns. I can do this with `awk`.

```{bash}
awk '{print NF}' 2019-06-20-mRNA-Descriptions.txt > 2019-06-20-mRNA-Description-Column-Numbers.txt
```

### Import isolated mRNA descriptions

```{r}
mRNADescriptionColumns <- read.delim("2019-06-20-mRNA-Description-Column-Numbers.txt", header = FALSE) #Import data with number of columns for each row
max(mRNADescriptionColumns) #Max number of columns. I'll need to import my gene description dataset with 78 columns per row.
```

```{r}
mRNADescriptions <- read.delim2("2019-06-20-mRNA-Descriptions.txt", header = FALSE, col.names = paste("V", 1:78, sep = "")) #Import document with gene descrptions. Name 78 columns so all rows are imported with 78 columns
head(mRNADescriptions) #Confirm import
```

```{r}
geneIDmRNAList <- mRNADescriptions$V3 #Isolate gene ID column
geneIDmRNAList <- gsub("Dbxref=GeneID:", "", geneIDmRNAList) #Remove extraneous characters
head(geneIDmRNAList) #Confirm changes
```

```{r}
genbankIDmRNAList <- mRNADescriptions$V4 #Isolate Genbank ID
genbankIDmRNAList <- gsub("Genbank:", "", genbankIDmRNAList) #Remove extraneous characters
head(genbankIDmRNAList) #Confirm changes
```

Buried in the `mRNAList` is mRNA product information. I'm going to extract that information and add it to the master list.

### Isolate product descriptions

```{bash}
cut -f9 ../2019-05-13-Generating-Genome-Feature-Tracks/C_virginica-3.0_Gnomon_mRNA_yrv.gff3 \
| awk  '{gsub("product=","\t",$0); print;}' \
| cut -f2 \
| awk  '{gsub(";","\t",$0); print;}' \
| awk  '{gsub("%","\t",$0); print;}' \
> 2019-06-20-mRNA-Products.txt #Isolate mRNA description column and convert "product=" to "\t". This separates the descriptions into 2 columns. The beginning of second column has the product description. Take the second column of the output and convert ";" and "%" to "\t" to better isolate mRNA products. Save the output as a new file.
```

```{bash}
head 2019-06-20-mRNA-Products.txt #Confirm product description isolation
```

```{bash}
awk '{print NF}' 2019-06-20-mRNA-Products.txt > 2019-06-20-mRNA-Products-Column-Numbers.txt
```

### Import isolated mRNA products

```{r}
mRNAProductColumns <- read.delim("2019-06-20-mRNA-Products-Column-Numbers.txt", header = FALSE) #Import data with number of columns for each row
max(mRNAProductColumns) #Max number of columns. I'll need to import my dataset with 16 columns per row.
```

```{r}
mRNAProducts <- read.delim2("2019-06-20-mRNA-Products.txt", header = FALSE, col.names = paste("V", 1:16, sep = "")) #Import document with gene descrptions. Name 16 columns so all rows are imported with 16 columns.
head(mRNAProducts) #Confirm import
```

```{r}
mRNAProducts <- mRNAProducts$V1 #Extract the first column with product information
```

### Merge mRNA track with `allDMLclosestBedgeneList`

```{r}
mRNAList <- data.frame(mRNAList, geneIDmRNAList, genbankIDmRNAList, mRNAProducts) #Replace mRNAList dataframe with mRNAList, geneIDmRNAList, genbankIDmRNAList, and mRNAProducts
mRNAList <- mRNAList[,-c(6:7)] #Remove extra columns
colnames(mRNAList) <- c("mRNA-chr", "Gnomon", "mRNA", "mRNA-start", "mRNA-end", "gene-ID", "Genbank", "Product") #Rename columns to match previous naming conventions
head(mRNAList) #Confirm merge
```

```{r}
allDMLclosestBedgeneListmRNAList <- merge(x = allDMLclosestBedgeneList, y = mRNAList, by = "gene-ID", sort = FALSE, all.x = TRUE) #Merge dataframes by gene ID
head(allDMLclosestBedgeneListmRNAList) #Confirm merge
```

```{r}
allDMLclosestBedgeneListmRNAList <- allDMLclosestBedgeneListmRNAList[,c(2:15, 1, 19:22)] #Reorganize columns
head(allDMLclosestBedgeneListmRNAList) #Confirm reorganization
```

```{r}
allDMLclosestBedgeneListmRNAList <- distinct(allDMLclosestBedgeneListmRNAList) #Keep only unique rows
```

## Merge with Uniprot codes

```{r}
allDMLclosestBedgeneListmRNAListUniprot <- merge(x = allDMLclosestBedgeneListmRNAList, y = blastResults, by = "Genbank", sort = FALSE, all.x = TRUE) #Merge DML list with Uniprot codes by Genbank IDs
head(allDMLclosestBedgeneListmRNAListUniprot) #Confirm merge
```

```{r}
allDMLclosestBedgeneListmRNAListUniprot <- allDMLclosestBedgeneListmRNAListUniprot[,c(2:4, 9, 6:8, 10:11, 14:18, 1, 21:22, 31, 19)] #Reorganize columns and save as a new dataframe
colnames(allDMLclosestBedgeneListmRNAListUniprot) <- c("chr", "start", "end", "meth.diff", "strand", "pvalue", "qvalue", "dist-to-closest-gene", "gene-chr", "gene-start", "gene-end", "geneID", "mRNA-start", "mRNA-end", "Genbank", "Uniprot-Accession", "Uniprot-ID", "evalue", "Product") #Rename columns
allDMLclosestBedgeneListmRNAListUniprot <- distinct(allDMLclosestBedgeneListmRNAListUniprot) #Keep only unique rows
head(allDMLclosestBedgeneListmRNAListUniprot) #Confirm reorganization
```

## Merge with GOterms

I will use a Uniprot-SwissProt database I downloaded to match Uniprot-Accession with Gene Ontology information, which includes GOterms and IDs. I downloaded a tab-delimited SwissProt database with five additional columns: Gene Ontology IDs, Gene ontology (GO), Gene ontology (biological processes), Gene ontology (cellular component), and Gene ontology (molecular function).

### Import Uniprot-Gene Ontology databse

```{bash}
gunzip uniprot-reviewed_yes.tab.gz #Unzip file
```

```{bash}
head -n2 uniprot-reviewed_yes.tab #View file header
```

```{r}
uniprotGOTerms <- read.delim("uniprot-reviewed_yes.tab", header = TRUE) #Import tab-delimited file with header
head(uniprotGOTerms) #Confirm import
```

```{r}
uniprotGOTerms <- uniprotGOTerms[,c(1:2, 8, 10:12)] #Remove unnecessary columns
colnames(uniprotGOTerms) <- c("Uniprot-Accession", "Uniprot-ID", "GO-ID", "GO-BP", "GO-CC", "GO-MF") #Rename columns to match previous naming conventions
head(uniprotGOTerms) #Confirm changes
```

### Merge with `allDMLclosestBedgeneListmRNAListUniprot`

```{r}
allDMLclosestBedgeneListmRNAListUniprotGOTerms <- merge(x = allDMLclosestBedgeneListmRNAListUniprot, y = uniprotGOTerms, by = c("Uniprot-Accession", "Uniprot-ID"), sort = FALSE, all.x = TRUE) #Merge files by Uniprot Accesssion code and Uniprot ID
head(allDMLclosestBedgeneListmRNAListUniprotGOTerms) #Confirm merge
```

```{r}
allDMLclosestBedgeneListmRNAListUniprotGOTerms <- allDMLclosestBedgeneListmRNAListUniprotGOTerms[,c(3:17, 1:2, 18:23)] #Reorganize columns
colnames(allDMLclosestBedgeneListmRNAListUniprotGOTerms) <- c("chr", "start", "end", "meth.diff", "strand", "pvalue", "qvalue", "dist-to-closest-gene", "gene-chr", "gene-start", "gene-end", "geneID", "mRNA-start", "mRNA-end", "Genbank", "Uniprot-Accession", "Uniprot-ID", "evalue", "Product", "GO-ID", "GO-BP", "GO-CC", "GO-MF") #Rename columns to match naming conventions
head(allDMLclosestBedgeneListmRNAListUniprotGOTerms) #Confirm changes
```

## Export master gene list

```{r}
masterDMLAnnotation <- distinct(allDMLclosestBedgeneListmRNAListUniprotGOTerms) #Remove duplicated rows (merging artifacts)
head(masterDMLAnnotation) #Confirm changes
```

```{r}
write.csv(masterDMLAnnotation, "2019-06-20-Master-DML-Annotation.csv") #Write out masterDMLAnnotation as a .csv file
```

Now that I have a master gene list, I can create relevant subset documents.

## DML-Gene overlaps

```{r}
DMLGeneAnnotation <- subset(x = masterDMLAnnotation, subset = masterDMLAnnotation$`dist-to-closest-gene` == 0) #Subset dataset where the distance to the closest gene = 0 (i.e. DML and gene are overlapping)
head(DMLGeneAnnotation) #Confirm subset
```

```{r}
write.csv(DMLGeneAnnotation, "2019-06-20-DML-Gene-Annotation.csv") #Save file
```

### Hypermethylated DML-gene overlaps

```{r}
DMLGeneAnnotationHypermethylated <- subset(x = DMLGeneAnnotation, subset = DMLGeneAnnotation$meth.diff > 0) #Subset hypermethylated DML by selecting meth.diff > 0
head(DMLGeneAnnotationHypermethylated) #Confirm subset
```

```{r}
write.csv(DMLGeneAnnotationHypermethylated, "2019-06-20-DML-Gene-Annotation-Hypermethylated.csv") #Save file
```

### Hypomethylated DML-gene overlaps

```{r}
DMLGeneAnnotationHypomethylated <- subset(x = DMLGeneAnnotation, subset = DMLGeneAnnotation$meth.diff < 0) #Subset hypomethylated DML by selecting meth.diff < 0
head(DMLGeneAnnotationHypomethylated) #Confirm subset
```

```{r}
write.csv(DMLGeneAnnotationHypomethylated, "2019-06-20-DML-Gene-Annotation-Hypomethylated.csv") #Save file
```

# Master exon overlap list

## Import DML-exon overlaps

```{r}
DMLExonOverlaps <- read.delim("../2018-11-01-DML-and-DMR-Analysis/2019-05-29-DML-Exon.txt", header = FALSE) #Import file with DML and matching exons
DMLExonOverlaps <- DMLExonOverlaps[,-4] #Remove extra meth.diff column
colnames(DMLExonOverlaps) <- c("chr", "start", "end", "exon-chr", "exon-start", "exon-end")
DMLExonOverlaps$start <- DMLExonOverlaps$start + 1 #Add 1 to the start position since the file is BEDtools output
DMLExonOverlaps$end <- DMLExonOverlaps$end - 1 #Add 1 to the start position since the file is BEDtools output
head(DMLExonOverlaps) #Confirm changes
```

## Merge with DML-gene annotation table

```{r}
DMLExonAnnotation <- merge(x = DMLExonOverlaps, y = DMLGeneAnnotation, by = "start", sort = FALSE, all.x = TRUE) #Merge files by DML position. Don't sort the output and retain all rows from DMLExonOverlaps even if there are no matches in DMLGeneAnnotation
head(DMLExonAnnotation) #Confirm merge
```

```{r}
DMLExonAnnotation <- DMLExonAnnotation[,c(2, 1, 3, 9:12, 5:6, 15:28)] #Reorganize columns remove dist-to-closest-gene column since these is all overlaps.
colnames(DMLExonAnnotation) <- c("chr", "start", "end", "meth.diff", "strand", "pvalue", "qvalue", "exon-start", "exon-end", "gene-start", "gene-end", "geneID", "mRNA-start", "mRNA-end", "Genbank", "Uniprot-ID", "Uniprot-Accession", "evalue", "Product", "GO-ID", "GO-BP", "GO-CC", "GO-MF") #Add column names
DMLExonAnnotation <- distinct(DMLExonAnnotation) #Remove duplicate rows (merging artifacts)
head(DMLExonAnnotation) #Confirm changes
```

```{r}
write.csv(DMLExonAnnotation, "2019-06-20-DML-Exon-Annotation.csv") #Save file
```

## Hypermethylated DML-exon overlaps

```{r}
DMLExonAnnotationHypermethylated <- subset(x = DMLExonAnnotation, y = DMLExonAnnotation$meth.diff > 0) #Obtain hypermethylated DML by subsetting meth.dif > 0
head(DMLExonAnnotationHypermethylated) #Confirm subset
```

```{r}
write.csv(DMLExonAnnotationHypermethylated, "2019-06-20-DML-Exon-Annotation-Hypermethylated.csv") #Save file
```

## Hypomethylated DML-exon overlaps

```{r}
DMLExonAnnotationHypomethylated <- subset(x = DMLExonAnnotation, y = DMLExonAnnotation$meth.diff < 0) #Obtain hypomethylated DML by subsetting meth.dif < 0
head(DMLExonAnnotationHypomethylated) #Confirm subset
```

```{r}
write.csv(DMLExonAnnotationHypomethylated, "2019-06-20-DML-Exon-Annotation-Hypomethylated.csv") #Save file
```

# Master intron overlap list

## Import DML-Intron overlaps

```{r}
DMLIntonOverlaps <- read.delim("../2018-11-01-DML-and-DMR-Analysis/2019-05-29-DML-Intron.txt", header = FALSE) #Import file with DML and matching introns
DMLIntonOverlaps <- DMLIntonOverlaps[,-4] #Remove extra meth.diff column
colnames(DMLIntonOverlaps) <- c("chr", "start", "end", "intron-chr", "intron-start", "intron-end")
DMLIntonOverlaps$start <- DMLIntonOverlaps$start + 1 #Add 1 to the start position since the file is BEDtools output
DMLIntonOverlaps$end <- DMLIntonOverlaps$end - 1 #Add 1 to the start position since the file is BEDtools output
head(DMLIntonOverlaps) #Confirm changes
```

## Merge with DML-gene annotation table

```{r}
DMLIntronAnnotation <- merge(x = DMLIntonOverlaps, y = DMLGeneAnnotation, by = "start", sort = FALSE, all.x = TRUE) #Merge files by DML position. Don't sort the output and retain all rows from DMLIntronOverlaps even if there are no matches in DMLGeneAnnotation
head(DMLIntronAnnotation) #Confirm merge
```

```{r}
DMLIntronAnnotation <- DMLIntronAnnotation[,c(2, 1, 3, 9:12, 5:6, 15:28)] #Reorganize columns remove dist-to-closest-gene column since these is all overlaps.
colnames(DMLIntronAnnotation) <- c("chr", "start", "end", "meth.diff", "strand", "pvalue", "qvalue", "intron-start", "intron-end", "gene-start", "gene-end", "geneID", "mRNA-start", "mRNA-end", "Genbank", "Uniprot-ID", "Uniprot-Accession", "evalue", "Product", "GO-ID", "GO-BP", "GO-CC", "GO-MF") #Add column names
DMLIntronAnnotation <- distinct(DMLIntronAnnotation) #Remove duplicate rows (merging artifacts)
head(DMLIntronAnnotation) #Confirm changes
```

```{r}
write.csv(DMLIntronAnnotation, "2019-06-20-DML-Intron-Annotation.csv") #Save file
```

## Hypermethylated DML-intron overlaps

```{r}
DMLIntronAnnotationHypermethylated <- subset(x = DMLIntronAnnotation, subset = DMLIntronAnnotation$meth.diff > 0) #Obtain hypermethylated DML by subsetting meth.diff > 0
head(DMLIntronAnnotationHypermethylated) #Confirm subset
```

```{r}
write.csv(DMLIntronAnnotationHypermethylated, "2019-06-20-DML-Intron-Annotation-Hypermethylated.csv") #Save file
```

## Hypomethylated DML-intron overlaps

```{r}
DMLIntronAnnotationHypomethylated <- subset(x = DMLIntronAnnotation, subset = DMLIntronAnnotation$meth.diff < 0) #Obtain hypomethylated DML by subsetting meth.diff < 0
head(DMLIntronAnnotationHypomethylated) #Confirm subset
```

```{r}
write.csv(DMLIntronAnnotationHypomethylated, "2019-06-20-DML-Intron-Annotation-Hypomethylated.csv") #Save file
```

# Master promoter overlap list

The `dist-to-closet-gene` has important information for DML distances to genes, but not the closest transcription start sites (i.e. mRNA). I will import the DML-promoter overlaps, then modify this dataframe to obtain a master DML-promoter overlap annotation.

## Import DML-promoter overlaps

```{r}
promoterList <- read.delim("../2018-11-01-DML-and-DMR-Analysis/2019-05-29-Flanking-Analysis/2019-05-29-mRNA-1000bp-UpstreamFlanks-DML.txt", header = FALSE) #Import DML-promoter overlaps
promoterList <- promoterList[,-c(7:8)] #Remove extra columns
colnames(promoterList) <- c("flank-chr", "flank-start", "flank-end", "chr", "start", "end") #Rename columns
head(promoterList) #Confirm changes
```

```{r}
promoterList$start <- promoterList$start + 1 #Add 1 to the start position since the file is BEDtools output
promoterList$end <- promoterList$end -1 #Subtract 1 to the end position since the file is BEDtools output
promoterList$`mRNA-start` <- promoterList$`flank-end` #Duplicate The end of the flank is the same as the start of the mRNA. Create a duplicate column for easy merging.
head(promoterList) #Confrm changes
```

## Merge with `masterDMLAnnotation`

```{r}
DMLPromoterAnnotation <- merge(x = masterDMLAnnotation, y = promoterList, by = "mRNA-start", sort = FALSE, all.y = TRUE) #Merge the master list with promoter information by mRNA-start. Retain all rows in promoterList that do not have matching entries in masterDMLAnnotation
head(DMLPromoterAnnotation) #Confirm merge
```

```{r}
DMLPromoterAnnotation <- DMLPromoterAnnotation[,c(2:13, 25:26, 1, 14:23)] #Reorganize columns
colnames(DMLPromoterAnnotation) <- c("chr", "start", "end", "meth.diff", "strand", "pvalue", "qvalue", "dist-to-closest-gene", "gene-chr", "gene-start", "gene-end", "geneID", "flank-start", "flank-end", "mRNA-start", "mRNA-end", "Genbank", "Uniprot-Accession", "Uniprot-ID", "evalue", "Product", "GO-ID", "GO-BP", "GO-CC", "GO-MF") #Rename columns to match naming conventions
DMLPromoterAnnotation <- distinct(DMLPromoterAnnotation) #Remove duplicate rows (merge artifacts)
head(DMLPromoterAnnotation) #Confirm reorganization
```

```{r}
write.csv(DMLPromoterAnnotation, "2019-06-20-DML-Promoter-Annotation.csv") #Save file
```

# Master transposable element list

The last annotation list I need is one with DML-transposable element overlaps.

## Import DML-TE overlap file

```{r}
TEOverlaps <- read.delim("../2018-11-01-DML-and-DMR-Analysis/2019-05-29-DML-TE-all.txt", header = FALSE) #Import DML-TE overlap file
TEOverlaps <- TEOverlaps[,-c(4, 6:7, 12)] #Remove extraneous columns
colnames(TEOverlaps) <- c("chr", "start", "end", "TE-chr", "TE-start", "TE-end", "V10", "TE-strand", "Target-Motif") #Rename columns
head(TEOverlaps) #Confirm import
```

## Merge with TE-gene overlap file

```{r}
TEGeneOverlaps <- read.delim("../2018-11-01-DML-and-DMR-Analysis/2018-11-07-Genes-TE-all.txt", header = FALSE)
TEGeneOverlaps <- TEGeneOverlaps[,-c(4:6, 11, 13)] #Remove extraneous columns
colnames(TEGeneOverlaps) <- c("gene-chr", "gene-start", "gene-end", "TE-start", "TE-end", "V10", "TE-strand", "Target-Motif") #Rename columns. Changed V9 to V10 to match column names in TEOverlaps
head(TEGeneOverlaps) #Confirm import
```

```{r}
TEDMLOverlaps <- merge(x = TEOverlaps, y = TEGeneOverlaps, by = c("TE-start", "TE-end"), sort = FALSE, all.x = TRUE) #Merge TEOverlaps with TEGeneOverlaps. Do not sort the output. Retain any information in the DML-TE overlaps that does not have matching gene annotations
head(TEDMLOverlaps) #Confirm merge
```

```{r}
TEDMLOverlaps <- TEDMLOverlaps[,c(3:5, 1:2, 7:9, 11:12)] #Reorganize columns
head(TEDMLOverlaps) #Confirm changes
```

## Merge with `masterDMLAnnotation`

This will allow me to match TE-DML-gene overlaps with genes.

```{r}
DMLTEAnnotation <- merge(x = TEDMLOverlaps, y = masterDMLAnnotation, by = c("gene-start", "gene-end"), sort = FALSE, all.x = TRUE) #Merge TEDMLOverlaps with masterDMLAnnotation by gene start and end positions. Do not sort the output and retain any TEDMLOverlap entries that do not have matches in masterDMLAnnotation
head(DMLTEAnnotation) #Confirm merge
```

```{r}
DMLTEAnnotation <- DMLTEAnnotation[,c(11:17, 6:10, 18, 1:2, 20:31)] #Reorganize columns
colnames(DMLTEAnnotation) <- c("chr", "start", "end", "meth.diff", "strand", "pvalue", "qvalue", "TE-start", "TE-end", "V10", "TE-strand", "Target-Motif", "dist-to-closest-gene", "gene-start", "gene-end", "geneID", "mRNA-start", "mRNA-end", "Genbank", "Uniprot-Accession", "Uniprot-ID", "evalue", "Product", "GO-ID", "GO-BP", "GO-CC", "GO-MF") #Rename columns
DMLTEAnnotation <- distinct(DMLTEAnnotation) #Retain unique rows
head(DMLTEAnnotation) #Confirm changes
```

```{r}
write.csv(DMLTEAnnotation, "2019-06-20-DML-TE-Annotation.csv") #Save file
```






















# Gene background

## Check file format

```{bash}
head 2018-12-18-Gene-Background-mRNA.txt #Check file format
```

The IDs in the blast result match up with the Genbank IDs in the overlap file. I need to isolate the Genbank IDs before I can merge the files. I'm not sure how to do this file manipulation in bash or R, so I'll do this in Excel.

```{bash}
head 2018-12-18-Gene-Background-mRNA-Unfolded.txt #Confirm unfolding
```

## Import files

```{r}
backgroundmRNA <- read.delim("2018-12-18-Gene-Background-mRNA-Unfolded.txt", header = FALSE) #Import background-mRNA overlaps
head(backgroundmRNA) #Confirm import
colnames(backgroundmRNA) <- c("backgroundchromosome", "backgroundstart", "backgroundend", "backgroundstrand", "V5","V6","V7","V8","V9","V10","V11","V12","V13","V14","V15","V16","V17","V18","V19","V20","V21","V22","V23","V24","V25","V26","V27","V28","V29","V30","V31","V32","V33","V34", "mRNAchromosome", "Gnomon", "mRNA", "mRNAstart", "mRNAend", "V40", "mRNAstrand", "V42", "ID", "Parent", "Dbxref=GeneID", "V46", "V47", "Genbank", "Name", "V50", "V51", "V52", "V53", "V54", "V55", "V56", "V57", "V58", "V59", "V60", "V61") #Rename columns
head(backgroundmRNA) #Confirm column naming
```

## Merge with Uniprot codes

```{r}
backgroundmRNAblast <- merge(x = blastResults, y = backgroundmRNA, by = "Genbank") #Merge the gene background and Uniprot codes by the Genbank ID column
head(backgroundmRNAblast) #Confirm merge
```

```{r}
write.csv(backgroundmRNAblast, "2019-02-22-mRNA-Gene-Background-Uniprot.csv") #Write out file
```

## Separate unique entries

I want a list of unique genes and matching Uniprot codes represented in the gene background-mRNA overlaps.

```{r}
uniqueBackgroundmRNAblast <- subset(backgroundmRNAblast, !duplicated(backgroundmRNAblast$Genbank)) #Subset the unique Genbank IDs from backgroundmRNAUniprot and save as a new dataframe.
head(uniqueBackgroundmRNAblast) #Confirm changes
```

```{r}
nrow(uniqueBackgroundmRNAblast) #Number of unique genes = 14943
write.csv(uniqueBackgroundmRNAblast, "2019-02-28-mRNA-Gene-Background-Uniprot-Unique.csv") #Save file
```

## Isolate Uniprot codes

To obtain gene enrichment information from DAVID, I need to isolate the Uniprot codes.

```{r}
backgroundmRNAUniprot <- backgroundmRNAblast$Uniprot #Isolate Uniprot codes from merged file
write.csv(backgroundmRNAUniprot, "2019-02-22-mRNA-Gene-Background-UniprotOnly.csv") #Save Uniprot codes as a new file
```

